A comparison on R tools for nonlinear least squares modeling

### 2023-03-05 RefactoringNLS/ComparingNLStools.Rmd
continue work

It is likely that there are a number of other tools that could be relevant to
this comparison in the Bioconductor (??ref) collection and on repositories like
Github (https://github.com) and Gitlab (https://about.gitlab.com), but we will 
not pursue those possibilities here.

@@

R tools are our interest, but the tools we consider often use other programming
languages. There are advantages in using a well-established program that has
been much used and tested. Early versions of R were relatively sluggish in 
computational speed. The downside is that we must interface between two or more
programming languages. Changes to one or the other language or its library support
force maintenance issues. Moreover, the pool of expertise with languages like C
and Fortran is diminishing.

@@

Early nonlinear least squares programs generally used the Gauss-Newton method 
(?? ref). This is often very effective and efficient, but for a quite large 
proportion of problems can fail to find the solution. Stabilization of the 
computations is relatively straightforward in principle, and overcomes most
of the failures, though sometimes at the cost of more calculations.

@@

@@

The suggestion is that $D$ be replaced by a multiple of the unit matrix or else
a multiple of the diagonal part of $J' J$. For machines with exceptionally low
precision arithmetic found on some minicomputers in the 1970s, one of us suggested
a linear combination of these ideas, as with low precision, some elements of 
$J' J$ could underflow to zero (@jn77ima). We note that the various choices for
$D$, especially the scaling multiple, as well as possible line search along the
direction $\delta$ rather than a unit step, give rise to a panoply of variant
algorithms. "Marquardt's method" is a family of methods. Fortunately, most choices
work well.

@@

Using a formula specification was one of the principal advances made
when `nls()` became available in S sometime in the 1980s. It uses a
Gauss-Newton (i.e., unstabilized) iteration with step reduction line
search. This works very efficiently as long as $J$ is not ill-conditioned.
In a one-parameter problem with ill-conditioning, think of $J$ as a 
very small number so that we need to watch for a zero-divide or 0/0 
situation in computing $\delta$.
The following show `nls()` does poorly on this example problem.

```{r ex02nls}
unls1<-try(nls(formula=frmu, start=stu1, data=weeddf))
summary(unls1)
snls1<-try(nls(formula=frms, start=sts1, data=weeddf))
summary(snls1)
tnls1<-try(nls(formula=frmt, start=stt1, data=weeddf))
summary(tnls1)
cat("\n")
```

Let us consider three alternatives to `nls()`: `nlsr::nlxb()`, `minpack.lm::nlsLM()`
and `gslnls::gsl_nls()`. We use default methods and controls in these examples.

@@

```{r ex02nlsr}
library(nlsr)
unlx1<-try(nlxb(formula=frmu, start=stu1, data=weeddf))
print(unlx1) 
pshort(unlx1) # A short form output
snlx1<-try(nlxb(formula=frms, start=sts1, data=weeddf))
# pshort(snlx1)
print(snlx1)
tnlx1<-try(nlxb(formula=frmt, start=stt1, data=weeddf))
# pshort(tnlx1)
print(tnlx1)
```

@@

```{r ex02minpack}
library(minpack.lm)
unlm1<-try(nlsLM(formula=frmu, start=stu1, data=weeddf))
summary(unlm1) # Use summary() to get display
unlm1 # Note the difference. Use this form to get sum of squares
# pnlm0(unlm1)  # Short form of output
snlm1<-try(nlsLM(formula=frms, start=sts1, data=weeddf))
# pnlm0(snlm1)
summary(snlm1)
## Following structure needed to deal with M1mac issue
tnlm1<-try(nlsLM(formula=frmt, start=stt1, data=weeddf))
if (inherits(tnlm1, "try-error")) {
   cat("Failure to compute solution -- likely singular Jacobian\n")
} else {  
   pnlm0(tnlm1) # short form to give sum of squares
   summary(tnlm1)
}   
```

@@

```{r ex02gslnls}
library(gslnls)
ugslnls1<-try(gsl_nls(fn = frmu, data = weeddf,  start = stu1))
summary(ugslnls1) # Use summary() to get display
sgslnls1<-try(gsl_nls(fn = frms, data = weeddf,  start = sts1))
summary(sgslnls1) # Use summary() to get display
tgslnls1<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1))
summary(tgslnls1) # Use summary() to get display
```

@@

?? what do we want here??

@@

Analytic Jacobian code can be provided to `nls()`, `nlsLM()` and `gsl_nls()`,
and most `selfStart` models that automatically provide starting parameters
also include such code. There is documentation in R
of `selfStart`, but the construction of them is non-trivial. Their use with
`nlsr` is described in the "Introduction to nlsr" vignette. However, since `nlsr`
generally can use very crude starting values, we have rarely needed them.

-------------------------------------------------

**nlsr** has functions to find a set of parameters that minimizes the sum of 
squares (or deviance) 
of differences between a model specified either as a formula using function 
`nlxb()` or as both residual
and Jacobian R functions using function `nlfb()`. As described below, the 
Jacobian may be approximated
by particular control settings.

Existing R tools that are similar to `nlxb()` are `nls()` and `minpack.lm::nlsLM()`;  
`minpack.lm::nls.lm()` is similar to `nlfb()`.

There are several important differences in approach and results, some of which 
are detailed in this document.

### Stabilization of Gauss-Newton computations

The iteration in all the major tools mentioned is the solution of the Gauss-Newton 
equations. `nls()` uses a variant of an approach suggested by @Hartley61, while 
`nlsr` and `minpack.lm` use variants of the method published by @Marquardt1963, 
though it was suggested earlier by @Levenberg1944. The details of the method in 
`nlsr` are discussed in  @JNABRefactoringNLS22.
Though it is unlikely to be useful in general, control settings for `nlxb()` or `nlfb()`
allow for those routines to perform a variety of Hartley and Marquardt algorithms. This
has been useful for learning about the algorithms, but tangential to simply finding
parameters of nonlinear models.

In general, the Levenberg-Marquardt stabilization is important in obtaining solutions.
The default method of `nls()` terminates with `singular gradient` unnecessarily in 
too many instances.

@@

An important choice made in developing `nlsr` was to code entirely within the R 
programming language. `nls()` uses a mix of R, C and Fortran, as does `minpack.lm`. 
Generally, I believe that keeping to a single programming language allows for 
easier maintenance and upgrades. On the other hand, R is usually somewhat slower 
in performing computations because it keeps track of names and structures and 
because it is usually interpreted rather than compiled. In recent years the 
performance penalty for using code entirely in R has been much reduced
with the just-in-time compiler and other improvements, so that using an all-R 
package offers acceptable performance. Indeed, in `nlsr` the use of R may be 
less of a performance cost than the attempt to ensure the solution has been found, 
which forces more iterations to be used.

@@

The code for `SSlogis()` is in `./src/library/stats/R/zzModels.R`.
<!-- ``` -->
<!-- SSlogis <- selfStart(~ Asym/(1 + exp((xmid - input)/scal)), -->
<!--     selfStart( -->
<!--         function(input, Asym, xmid, scal) -->
<!--         { -->
<!--               .expr1 <- xmid - input -->
<!--               .expr3 <- exp(.e2 <- .expr1/scal) -->
<!--               .expr4 <- 1 + .expr3 -->
<!--               .value <- Asym/.expr4 -->
<!--               .actualArgs <- as.list(match.call()[c("Asym", "xmid", "scal")]) -->
<!--               if(all(vapply(.actualArgs, is.name, NA))) -->
<!--               { -->
<!--             		  .expr10 <- .expr4^2 -->
<!--                   .grad <- array(0, c(length(.value), 3L), list(NULL, c("Asym", "xmid", "scal"))) -->
<!--                   .grad[, "Asym"] <- 1/.expr4 -->
<!-- 		              .grad[, "xmid"] <- - (xm <- Asym * .expr3/scal/.expr10) -->
<!-- 		              .grad[, "scal"] <- xm * .e2 -->
<!--                   dimnames(.grad) <- list(NULL, .actualArgs) -->
<!--                   attr(.value, "gradient") <- .grad -->
<!--               } -->
<!--               .value -->
<!--         }, -->
<!--         initial = function(mCall, data, LHS, ...) { -->
<!--               xy <- sortedXyData(mCall[["input"]], LHS, data) -->
<!--               if(nrow(xy) < 4) { -->
<!--                   stop("too few distinct input values to fit a logistic model") -->
<!--               } -->
<!--               z <- xy[["y"]] -->
<!--               ## transform to proportion, i.e. in (0,1) : -->
<!--               rng <- range(z); dz <- diff(rng) -->
<!--               z <- (z - rng[1L] + 0.05 * dz)/(1.1 * dz) -->
<!--               xy[["z"]] <- log(z/(1 - z))		# logit transformation -->
<!--               aux <- coef(lm(x ~ z, xy)) -->
<!--               pars <- coef(nls(y ~ 1/(1 + exp((xmid - x)/scal)), -->
<!--                                data = xy, -->
<!--                                start = list(xmid = aux[[1L]], scal = aux[[2L]]), -->
<!--                                algorithm = "plinear", ...)) -->
<!--               setNames(pars [c(".lin", "xmid", "scal")], -->
<!--                        mCall[c("Asym", "xmid", "scal")]) -->
<!--         }, -->
<!--         parameters = c("Asym", "xmid", "scal")) -->
<!-- ``` -->
This R function includes analytic expressions for the Jacobian ("gradient").
These could be useful to R users, especially if documented. Moreover, we
wonder why the programmers have chosen to save so many quantities in "hidden"
variables, i.e., with names preceded by ".". These are then not displayed by the `ls()`
command, making them difficult to access.

@@

<!-- ### Example output -->

<!-- To illustrate, let us run the Croucher example. -->

<!-- ```{r nlsoutx} -->
<!-- # Croucher example -->
<!-- xdata <- c(-2,-1.64,-1.33,-0.7,0,0.45,1.2,1.64,2.32,2.9) -->
<!-- ydata <- c(0.699369,0.700462,0.695354,1.03905,1.97389,2.41143,1.91091,0.919576,-0.730975,-1.42001) -->
<!-- p1<- 1; p2<-0.2; NLSstart<-list(p1=p1,p2=p2) -->
<!-- NLSformula <- ydata ~ p1*cos(p2*xdata) + p2*sin(p1*xdata) -->
<!-- NLSdata<-data.frame(xdata, ydata) -->
<!-- # Try full output version of nls -->
<!-- library(nlspkg) # use the packaged version of nls() -->
<!-- result<-nls(NLSformula, data=NLSdata, start=NLSstart, model=TRUE) -->
<!-- # We can display the result with several commands -->
<!-- # str(result) -- displays large amount of material - suppressed here -->
<!-- #    as it is too wide for the page -->
<!-- # result # displays the object -->
<!-- # ls(result) # to list the elements of the output -->
<!-- # ls(result$m) # and in particular the "m" object -->
<!-- ``` -->

<!-- ### Concerns with content of the nls result object -->

@@

```{r chkbound}
# bhobbsX.R ## bounded formula specification of problem using Hobbs Weed problem
weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
          38.558, 50.156, 62.948, 75.995, 91.972)
tt <- 1:12
wf <- data.frame(y = weed, tt = tt)
st <- c(b1 = 1, b2 = 1, b3 = 1) # a default starting vector (named!)
wmods <- y ~ 100 * b1 / (1 + 10 * b2 * exp(-0.1 * b3 * tt)) ## Scaled model
require(minpack.lm)
# Hobbs scaled problem with bounds, formula specification
anlxb1 <- nlxb(wmods, start = st, data = wf, lower = c(0, 0, 0), upper = c(2, 6, 3))
cat("Using nlsr::nlxb():")
print(anlxb1)
## nlsLM seems NOT to work properly with bounds
anlsLM1b <- nlsLM(wmods, start = st, data = wf, lower = c(0, 0, 0), upper = c(2, 6, 3))
cat("\n"); cat("using minpack.lm::nlsLM():")
print(anlsLM1b)
anlsb<-nls(wmods, start = st, data = wf, algorithm = "port", 
           lower = c(0, 0, 0), upper = c(2, 6, 3))
cat("Using nls with 'port':")
print(anlsb)
```

@@

<!-- For `nls()` the concern is demonstrated by a short email query from John Nash -->
<!-- to Doug Bates and his reply. -->
<!-- This is NOT a criticism of Prof. Bates' (or any other) work, but a reflection on  -->
<!-- how difficult it is to develop code in this subject area and to keep it maintainable.  -->
<!-- We have experienced similar loss of understanding for some of our own codes. -->
<!--     ... -->
<!--     https://gitlab.com/nashjc/improvenls/-/blob/master/Croucher-expandednlsnoc.R -->
<!--     ... has the test problem and the expanded code. Around line 367 is where we are -->
<!--     scratching our heads. The function code (from nlsModel()) is in the commented  -->
<!--     lines below the call. This is -->
<!--           # > setPars -->
<!--           # function(newPars) { -->
<!--           #   setPars(newPars) -->
<!--           #   resid <<- .swts * (lhs - (rhs <<- getRHS())) # envir = thisEnv {2 x} -->
<!--           #   dev   <<- sum(resid^2) # envir = thisEnv -->
<!--           #   if(length(gr <- attr(rhs, "gradient")) == 1L) gr <- c(gr) -->
<!--           #   QR <<- qr(.swts * gr) # envir = thisEnv -->
<!--           #   (QR$rank < min(dim(QR$qr))) # to catch the singular gradient matrix -->
<!--           # } -->
<!--     I'm anticipating that we will be able to set up a (possibly inefficient) code -->
<!--     with documentation that will be easier to follow and test, then gradually figure -->
<!--     out how to make it more efficient. -->
<!--     The equivalent from minpack.lm is -->
<!--     setPars = function(newPars) { -->
<!--                 setPars(newPars) -->
<!--                 assign("resid", .swts * (lhs - assign("rhs", getRHS(), -->
<!--                     envir = thisEnv)), envir = thisEnv) -->
<!--                 assign("dev", sum(resid^2), envir = thisEnv) -->
<!--                 assign("QR", qr(.swts * attr(rhs, "gradient")), envir = thisEnv) -->
<!--                 return(QR$rank < min(dim(QR$qr))) -->
<!--             } -->
<!--     In both there is the recursive call, which must have a purpose I don't understand. -->
<!-- The reply was -->
<!--     I'm afraid that I don't know the purpose of the recursive call either.   -->
<!--     I know that I wrote the code to use a closure for the response, covariates, etc.,  -->
<!--     but I don't recall anything like a recursive call being necessary. -->
<!--      I don't think I will be of much help.  My R skills have atrophied to the point  --> 
<!--      where I wouldn't even know how to start exploring what is happening in the first  -->
<!--      call as opposed to the recursive call. --> 
<!--  ``` -->

@@

<!-- Arkajyoti -- do you want to expand? -->
<!-- No. I think it would be tangential to the goal of the paper. We can have a -->
<!-- separate article for that, maybe? -->
<!-- Trying to trim a lot of words, so am making reference to package. Is it ready -->
<!-- for others to use? -->

<!-- When we have a "new" or trial solver function, we would like to know if it gives  -->
<!-- acceptable results on a range of sample problems of different types, starting -->
<!-- parameters, input conditions, constraints, subsets, weights or other settings.  -->
<!-- Ideally we want to be able to get a summary that is easy to read and assess. For -->
<!-- example, one approach would be to list the names of a set of tests with a red, green -->
<!-- or yellow dot beside the name for FAILURE, SUCCESS, or "NOT APPLICABLE". In the last -->
<!-- category would be a problem with constraints that the solver is not designed to  -->
<!-- handle.  -->

<!-- To accomplish this, we need a suitable "runner" program that can be supplied with -->
<!-- the name of a solver or solvers and a list of test problem cases. Problems generally -->
<!-- have a base setup -- a specification of the function to fit as a formula, some data -->
<!-- and a default starting set of parameters. Other cases can be created by imposing -->
<!-- bounds or mask constraints, subsets of the data, and different starts.  -->

<!-- How to set up this "runner" and its supporting infrastructure is non-trivial. While -->
<!-- the pieces are not as complicated as the inter-related parts of the solvers, especially -->
<!-- `nls()`, the categorization of tests, their documentation, and the structuring to make -->
<!-- running them straightforward and easy requires much attention to detail. -->

<!-- Some considerations for our test scripts: -->

<!-- - Is it useful to have a "base" script for each family of test problem, with numbered particular -->
<!--   cases? That is, if we run the scripts in order, we can avoid some duplication of code and -->
<!--   data. -->

<!-- - While we have developed some tags to document the test problem families and cases, we believe -->
<!--   that such tags (essentially summary documentation) will continue to need revision as different -->
<!--   tools and problems are included in scope of `nlsCompare`. -->

<!-- - Similarly, we expect that there will be ongoing review of the structure of the result files. -->

<!-- # Progress so far -->

<!-- Much of our work is reported in https://github.com/nashjc/RNonlinearLS . -->

<!-- ## Reports and articles -->

<!-- - The present article is the central report of the Google Summer or Code 2021 project -->
<!--   "Improvements to nls()". -->

<!-- - VarietyInNonlinearLeastSquaresCodes.Rmd: a review of the different algorithms and -->
<!--   the many choices in their implementation for nonlinear least squares. This is still -->
<!--   a DRAFT at 2022-6-20. -->

<!-- Response to query about TestsDoc -->
<!-- 4. Testing infrastructure, I think, is a general topic and not -->
<!-- particularly relevant to nls(); i.e., it is important for all packages. -->
<!-- So, I don't think we need to include that in RefactoringNLS. Could be a -->
<!-- separate blog maybe?   -->

<!-- Arkajyoti: I think TestsDoc is too preliminary to mention, though the links might be -->
<!-- useful. Perhaps incorporate in the nlsCompare repo. -->
<!-- - TestsDoc.Rmd (?citation) is a survey of testing tools in R. It has more  -->
<!--   general possibilities  -->
<!--   and fits into the subject of regression testing, in which case a  -->
<!--   more extensive literature review will be needed.  Note that this -->
<!--   document reflects the work in the the "Problem sets and test infrastructure" -->
<!--   below. -->

  <!-- need to incorporate MachID and -->
  <!-- the many tests. Arkajyoti: should we think of working this up into a paper for -->
  <!-- JSS or the R-Journal? My  -->
  <!-- view is that this COULD be a long term side-interest for your academic work, but -->
  <!-- that would depend on your own interests as well as opportunities. -->

<!-- This needs more focus of where it may appear. I'm thinking of a blog post -- JN-->
<!-- - DerivsNLS.Rmd ?citation: a document to explain different ways in which Jacobian information -->
<!--   is supplied to nonlinear least squares computation in R. File `ExDerivs.R` is -->
<!--   a DRAFT of a script to provide examples. -->

@@

<!-- ?? probably don't need to include -->
<!-- - WorkingDocument4ImproveNLS.Rmd: essentially a way to record what we have worked on. -->
<!--   A project diary. -->

<!-- ?? does not seem to add enough? -->
<!-- ## Problem sets and test infrastructure -->

<!-- ?? need to clean up -->

<!-- We have several test problems and variants thereof in the `inst/scripts/` directory of -->
<!-- the `nlsCompare` package available on Github (https://github.com/ArkaB-DS/nlsCompare). -->
<!-- We direct the reader to that package for documentation of the test infrastructure,  -->
<!-- in particular the problems and methods files (`problems.csv` and `methods.csv`) and -->
<!-- the various functions invoked by `run.R` to produce an output file in CSV form also. -->

<!-- Towards the end of the project, we have focused our attention on the `nlsCompare` -->
<!-- package, which looks at evaluating and comparing functions for nonlinear least -->
<!-- squares problems. We use many of the same tests as checks that are or will be  -->
<!-- built into our packages for such problems e.g., `nlsalt`. These use the `testthat` -->
<!-- structure and may include verification of outputs that are specific to a package, -->
<!-- such as the upper triangular matrix R of the QR decomposition that has been  -->
<!-- computed for a Jacobian. Since such an object will not be computed by all methods,  -->
<!-- testing it in `nlsCompare` makes no sense, and in that package we concentrate on -->
<!-- the minimum sum of squares and the associated model parameters. -->

<!-- ## Code development -->

<!-- A summary of our main results: -->

<!-- - nlspkg: a packaged version of the `nls() code from R-base. Thanks to Duncan Murdoch -->

<!-- - nlsalt: attempt to mirror `nls() behaviour entirely in R. This is UNFINISHED.  -->
<!--   The effort showed that the structure of the programs was difficult to follow,  -->
<!--   undocumented, and unsuited to adding improvements like the Marquardt stabilization.  -->
<!--   We were able to get a pure-R version of `numericDeriv()` and rework most of the functions  -->
<!--   of `nlsModel` (but not `nlsModel.plinear`). This work may continue after the project -->
<!--   formally ends, but collaboration is likely needed with workers who have a deep knowledge -->
<!--   of R internals. `nls-flowchart.txt` was a start at documentation of the structure of `nls()`. -->

<!-- <!-- - nls-changes-for-small-residuals-in-nls-R-4.0.2.zip: collected material for the  --> 
<!-- <!--   fix by JN to the relative offset convergence criterion failure when there are small  --> 
<!-- <!--   residuals in problems sent to ``nls()`. Now part of R. -->

<!-- - the safeguarded relative offset convergence criterion that allows small-residual problems -->
<!--   to ba handled with `nls()` is now in the R base code. -->

<!-- - `nlsj` is a refactoring of some `nls()` functionality. ?? What are key features?? -->
<!--   This is an interim package for exploration of ideas and will NOT, as it currently -->
<!--   is presented, become a distributed package. -->

<!-- - nlsralt: this is intended to be a modified version of Nash and Murdoch package  -->
<!--   `nlsr` to examine possible improvements discovered as a result of this project.  -->
<!--   This is EXPERIMENTAL. -->

<!-- For example,  -->
<!-- `nlsr` does not currently use the `subset` argument correctly. We also need to document -->
<!-- changes e.g., to numerical derivatives. However, these are tasks for JN rather than AB. -->

>>>>

It is likely that there are a number of other tools that could be relevant to
this comparison in the Bioconductor (@Bioconductor) collection 
and on repositories like
Github (https://github.com) and Gitlab (https://about.gitlab.com), but we will 
not pursue those possibilities here.

@@

R tools are our interest, but they often call programs in other programming
languages. There are advantages in using a well-established program that has
been much used and tested. Early versions of R were relatively sluggish in 
computational speed. The downside is that we must interface between two or more
programming languages, adding to the risk of errors. Changes to one or the 
other language or its library support may
force maintenance issues. Moreover, the pool of expertise with languages like C
and Fortran is diminishing.

@@

Early nonlinear least squares programs generally used the Gauss-Newton method 
(@Hartley1961). This is often very effective and efficient, but for a quite large 
proportion of problems can fail to find the solution. Stabilization of the 
computations is relatively straightforward in principle, and overcomes most
of the failures, though sometimes at the cost of more calculations.

@@

### Documentation

?? presence of vignettes, demos, (good?) examples

@@

The suggestion is that $D$ be replaced by a multiple of the unit matrix or else
a multiple of the diagonal part of $J' J$. For machines with exceptionally low
precision arithmetic found on some minicomputers in the 1970s, one of us suggested
a linear combination of these ideas, as with low precision, some elements of 
$J' J$ could underflow to zero (@jn77ima). We note that the various choices for
$D$, especially the scaling multiple, as well as possible line search along the
direction $\delta$ rather than a unit step (@Hartley1961), 
give rise to a panoply of variant
algorithms. "Marquardt's method" is a family of methods. Fortunately, most choices
work well.

@@

Using a formula specification was one of the principal advances made
when `nls()` became available in S sometime in the 1980s. It uses a
Gauss-Newton (i.e., unstabilized) iteration with step reduction line
search. This works very efficiently as long as $J$ is not ill-conditioned.
In a one-parameter problem with ill-conditioning, think of $J$ as a 
very small number so that we need to watch for a zero-divide or 0/0 
situation in computing $\delta$.
The following show `nls()` does poorly on this example problem. Note that `nlsr`
offers function `pnlm0()` to give a 1-line display of class nls method results,
and `pshort()` for `nlsr`.

```{r shortforms, echo=FALSE}
library(nlsr)
```

```{r ex02nls}
unls1<-try(nls(formula=frmu, start=stu1, data=weeddf))
summary(unls1)
snls1<-try(nls(formula=frms, start=sts1, data=weeddf))
summary(snls1)
tnls1<-try(nls(formula=frmt, start=stt1, data=weeddf))
summary(tnls1)
```

Let us consider three alternatives to `nls()`: `nlsr::nlxb()`, `minpack.lm::nlsLM()`
and `gslnls::gsl_nls()`, which we will call the **class nls** methods.
We use default methods and controls in these examples.

@@

```{r ex02nlsr}
library(nlsr)
unlx1<-try(nlxb(formula=frmu, start=stu1, data=weeddf))
print(unlx1) 
pshort(unlx1) # A short form output
snlx1<-try(nlxb(formula=frms, start=sts1, data=weeddf))
pshort(snlx1) # or print(snlx1)
tnlx1<-try(nlxb(formula=frmt, start=stt1, data=weeddf))
pshort(tnlx1) # alternatively print(tnlx1)
```

@@

```{r ex02minpack}
library(minpack.lm)
unlm1<-try(nlsLM(formula=frmu, start=stu1, data=weeddf))
summary(unlm1) # Use summary() to get display
# unlm1 # Note the difference. Use this form to get sum of squares
# pnlm0(unlm1)  # Short form of output
snlm1<-try(nlsLM(formula=frms, start=sts1, data=weeddf))
pnlm0(snlm1)
# summary(snlm1)
## Following structure needed to deal with M1mac issue
tnlm1<-try(nlsLM(formula=frmt, start=stt1, data=weeddf))
if (inherits(tnlm1, "try-error")) {
   cat("Failure to compute solution -- likely singular Jacobian\n")
} else {  
   pnlm0(tnlm1) # short form to give sum of squares, else use summary(tnlm1)
}   
```

@@

```{r ex02gslnls}
library(gslnls)
ugslnls1<-try(gsl_nls(fn = frmu, data = weeddf,  start = stu1))
summary(ugslnls1) # Use summary() to get display
pnlm0(ugslnls1) # to get sum of squares
sgslnls1<-try(gsl_nls(fn = frms, data = weeddf,  start = sts1))
pnlm0(sgslnls1) # Use summary() to get display
tgslnls1<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1))
pnlm0(tgslnls1) 
```

@@

We illustrate how to solve nonlinear least squares problems using
a function to define the residual. Note that `gsl_nls()` requires
a vector `y` that is the expected value of what we have called the
residual, but in this case is actually the model. `gsl_nls` uses 
a numerical approximation for the Jacobian if the argument `jac` 
is missing.

```{r exhobbsfn, eval=TRUE}
require(minpack.lm)
require(nlsr)
require(gslnls)
hobnlfb<-nlfb(start=stu1, resfn=hobbs.res, jacfn=hobbs.jac)
print(hobnlfb)
hobnlm<-nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac)
summary(hobnlm)  
hobgsln<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12))
summary(hobgsln)
hobgsl<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12),jac=hobbs.jac)
summary(hobgsl)
```

@@

Analytic Jacobian code can be provided to `nls()`, `nlsLM()` and `gsl_nls()`,
and most `selfStart` models that automatically provide starting parameters
also include such code. There is documentation in R
of `selfStart` models, but the construction of them is non-trivial. Their use with
`nlsr` is described in the "Introduction to nlsr" vignette. However, since `nlsr`
generally can use very crude starting values, we have rarely needed them.

## Bounds constraints on parameters

In some cases, we know that parameters cannot be bigger or smaller than some
externally known limits. Cash reserves cannot be negative. Animals have a 
minimum need for water. Airplanes cannot carry more than a known or 
legislated cargo weight. Such limits can be built into models, but there 
are some important details for using the tools in R.

- `nls()` can only impose bounds if the `algorithm="port"` argument is 
used in the call. Unfortunately, the documentation warns us:

  *The algorithm = "port" code appears unfinished, and does not even 
   check that the starting value is within the bounds. Use with 
   caution, especially where bounds are supplied.*

- `gsl_nls()` does not offer bounds.

- `nlsLM()` includes bounds in the standard call, but we have observed cases
   where it fails to get the correct answer. From examination of the code,
   we believe the authors have not taken into account all possibilities, though
   it should be noted that all programs have some weakness in
   regard to constrained optimization. Software creators have to work with
   assumptions on the extremity of scale that they are willing to countenance,
   and sometimes problems will be outside the scope envisaged.

```{r ex10, echo=TRUE} 
# Start MUST be feasible i.e. on or within bounds
anlshob1b <- nls(frms, start=sts1, data=weeddf, lower=c(0,0,0),
             upper=c(2,6,3), algorithm='port')
pnls0(anlshob1b) #  check the answer (short form)
# nlsLM seems NOT to work with bounds in this example
anlsLM1b <- nlsLM(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
pnls0(anlsLM1b)
# also no warning if starting out of bounds, but gets good answer!!
st4<-c(c1=4, c2=4, c3=4)
anlsLMob <- nlsLM(frms, start=st4, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
pnls0(anlsLMob)
# Try nlsr::nlxb()
anlx1b <- nlxb(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
pshort(anlx1b)
```

Let us try to fix (mask) the first parameter in the first two example problems.

```{r ex10m, echo=TRUE}
# Hobbsmaskx.R -- masks with formula specification of problem
require(nlsr)
require(minpack.lm)
stu <- c(b1=200, b2=50, b3=0.3) # a default starting vector (named!)
sts <- c(c1=2, c2=5, c3=3) # a default scaled starting vector (named!)
traceval<-FALSE
# fix first parameter
anxbmsk1 <- try(nlxb(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
			upper=c(200, 60, .3), trace=traceval))
print(anxbmsk1)
anlM1 <- try(nlsLM(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
			upper=c(200, 60, .3), trace=traceval))
summary(anlM1)
# nls gives warnings
anlsmsk1 <- try(nls(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
		upper=c(200, 60, .3),  algorithm="port", trace=traceval))
summary(anlsmsk1)
# Hobbs scaled problem with bounds, formula specification
anlxmsks1 <- nlxb(frms, start=sts, data=weeddf, lower=c(2,0,0),
                  upper=c(2,6,3))
print(anlxmsks1)
anlshmsk1 <- nls(frms, start=sts, trace=traceval, data=weeddf, lower=c(2,0,0),
             upper=c(2,6,3), algorithm='port')
summary(anlshmsk1)
cat("More precisely...crossprod(resid(anlshmsk1))=",crossprod(resid(anlshmsk1)),"\n")
# nlsLM does not always work with bounds
anlsLMmsks1 <- nlsLM(frms, start=sts, data=weeddf, lower=c(2,0,0),
                 upper=c(2,6,3))
summary(anlsLMmsks1)

# Test with all parameters masked
anlxmskall<- try(nlxb(frms, start=sts, data=weeddf, lower=sts, upper=sts))
print(anlxmskall)
```

`nlsr` has an output format that indicates the constraint status of the parameter
estimates. For `nlsr` we have **chosen** to suppress calculation of 
approximate standard 
errors in the parameters when constraints are active, though we believe this 
worthy of discussion and further investigation.

## Stabilization of Gauss-Newton computations

All four major mentioned must solve some variant of the Gauss-Newton 
equations. `nls()` uses a modification of an approach suggested by @Hartley1961,
while `nlsr`, `gslnls` and `minpack.lm` use flavours of @Marquardt1963. `gslnls`
offers an accelerated Marquardt method and three alternative methods. We have 
not explored all five methods with various control settings.
Though it is unlikely to be useful in general, 
control settings for `nlxb()` or `nlfb()` allow for those routines to perform 
a variety of Hartley and Marquardt algorithms. This
has been useful for learning about the algorithms, but tangential to simply finding
parameters of nonlinear models.

In general, the Levenberg-Marquardt stabilization is important in obtaining 
solutions in methods of the Gauss-Newton family, as `nls()` terminates 
too frequently and unnecessarily with `singular gradient` errors.

@@

An important choice made in developing `nlsr` was to code entirely within the R 
programming language. `nls()` uses a mix of R, C and Fortran, as does `minpack.lm`. 
As far as we are aware, gslnls
Generally, I believe that keeping to a single programming language allows for 
easier maintenance and upgrades. On the other hand, R is usually somewhat slower 
in performing computations because it keeps track of names and structures and 
because it is usually interpreted rather than compiled. In recent years the 
performance penalty for using code entirely in R has been much reduced
with the just-in-time compiler and other improvements, so that using an all-R 
package offers acceptable performance. Indeed, in `nlsr` the use of R may be 
less of a performance cost than the attempt to ensure the solution has been found, 
which forces more iterations to be used.

@@

The code for `SSlogis()` is in `./src/library/stats/R/zzModels.R`.

This R function includes analytic expressions for the Jacobian ("gradient").
These could be useful to R users, especially if documented. Moreover, we
wonder why the programmers have chosen to save so many quantities in "hidden"
variables, i.e., with names preceded by ".". These are then not displayed by the `ls()`
command, making them difficult to access.

@@

@@

@@

@@

@@
